---
title: "RePsychLing Barr & Seyfeddinipur (2011)"
author: "Reinhold Kliegl"
date: "05 March, 2015"
output:
  html_document: default
  pdf_document:
    highlight: tango
    keep_tex: yes
  word_document: default
geometry: margin=1in
fontsize: 12pt
bibliography: RePsychLing.bib
---
<!-- 
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{RePsychLing Barr & Seyfeddinipur (2011)}
-->

```{r preliminaries,echo=FALSE,include=FALSE,cache=FALSE}
library(lme4)
library(RePsychLing)
library(knitr)
options(width=92)
```

## Data from @barrseyfedd2010

Some of the data from @barrseyfedd2010 are available as the data frame `bs10` in the `RePsychLing` package.  

```{r bs10str}
str(bs10)
```

As with other data frames in this package, the subject and item factors are called `subj` and `item`.  The response being modelled, `dif`, variable is the difference in two response times.

The two experimental factors `S` and `F`, both at two levels, are represented in the -1/+1 encoding, as is their interaction, `SF`.  The `S` factor is the speaker condition with levels -1 for the same speaker in both trials and +1 for different speakers.  The `F` factor is the filler condition with levels -1 for `NS` and +1 for `FP`.

### Maximal linear mixed model (_maxLMM_)

The maximal model has a full factorial design `1+S+F+SF` for the fixed-effects and for potentially correlated vector-valued random effects for the subject and the item.

```{r m0}
print(summary(m0 <- lmer(dif ~ 1+S+F+SF + (1+S+F+SF|subj) + (1+S+F+SF|item), bs10, REML=FALSE)), corr=FALSE)
```

The model converges with warnings. Nine of 12 correlation parameters are estimated at the +/-1 boundary. 

Notice that there are only 12 items.  Expecting to estimate 4 variances and 6 covariances from 12 items is optimistic.

### PCA analysis of the _maxLMM_

A summary of a principal components analysis (PCA) of the random-effects variance-covariance
matrices shows
```{r sv_m0_s}
summary(rePCA(m0))
```

For both the by-subject and the by-item random effects the estimated variance-covariance matrices are singular.  There are at least 2 dimensions with no variation in the subject-related random-effects and 3 directions with no variation for the item-related random-effects.

Clearly the model is over-specified.

### Zero-correlation-parameter linear mixed model (_zcpLMM_)

A zero-correlation-parameter model fits independent random effects for the intercept, the experimental factors and their interaction for each of the `subj` and `item` grouping factors.
It can be conveniently specified using the `||` operator in the random-effects terms.
```{r m1}
print(summary(m1 <- lmer(dif ~ 1+S+F+SF + (1+S+F+SF||subj) + (1+S+F+SF||item), bs10, REML=FALSE)), corr=FALSE)
anova(m1, m0)
```

The _zcpLMM_ fits significantly worse than the _maxLMM_. However, the results strongly suggest that quite a few of the variance components are not supported by the data. 

### PCA for the _zcpLMM_


```{r pca1}
summary(rePCA(m1))
```

The random effect for filler, `F`, by subject has essentially zero variance and the random effect for the interaction, `SF`, accounts for less than 15% of the total variation in the random effects.  There is no evidence for item-related random effects in `m1`.

### Iterative reduction of model complexity

Remove all variance components estimated with a value of zero. 

```{r m2}
print(summary(m2 <- lmer(dif ~ 1+S+F+SF +  (1+S+SF||subj), bs10,  REML=FALSE)), corr=FALSE)
```

Naturally, the fit for this model is equivalent to that for `m1` because it is only the variance components with zero estimates that are eliminated.
```{r m21anova}
anova(m2,m1)
```

Next we check whether the variance component for the interactionm, `SF`, could reasonably be zero.

```{r m3}
print(summary(m3 <- lmer(dif ~ 1+S+F+SF +  (1+S||subj), bs10, REML=FALSE)), corr=FALSE)
anova(m3, m2)
```

Not quite significant, but could be considered. The fit is still worse than for the _maxLMM_ `m0`. We now reintroduce a correlation parameters in the vector-valued random effects for `subj`.

### Extending the reduced LMM with correlation parameters

```{r m4}
print(summary(m4 <- lmer(dif ~ 1+S+F+SF + (1+S|subj), bs10,  REML=FALSE)), corr=FALSE)
anova(m3, m4, m0)
```

Looks like we have evidence for a significant correlation parameter. Moreover, LMM `m4` fits as well as _maxLMM_. 

### Summary

LMM `m4` is the optimal model. It might be worth while to check the theoretical contribution of the correlation parameter.  

## Versions of packages used
```{r versions}
sessionInfo()
```

# References

