---
title: "Revisiting Gann & Barr (in press)"
author: "Reinhold Kliegl"
date: "06 May, 2015"
output:
  html_document: default
  pdf_document:
    highlight: tango
    keep_tex: yes
  word_document: default
geometry: margin=1in
fontsize: 12pt
---
<!-- 
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Revisiting Gann & Barr (2012)}
-->

```{r preliminaries,echo=FALSE,include=FALSE,cache=FALSE}
library(RePsychLing)
library(lme4)
library(knitr)
options(width=92,show.signif.stars = FALSE)
```

## Data from Gann & Barr (2012)

These data are available as `gb12` in the `RePsychLing` package.

```{r gbstr}
str(gb12)
summary(gb12)
```

### Maximal linear mixed model (_maxLMM_)

We assume `P`, the partner, is a between-session factor and `F`, feedback, is a between-item factor (i.e., they are not included in RE terms). The model fit in the paper is:

```{r m0}
print(summary(m0 <- lmer(
    sottrunc2 ~ 1+T+P+F+TP+TF+PF+TPF+(1+T+F+TF|session)+(1+T+P+TP|item), gb12,REML=FALSE)), corr=FALSE)
```

The model converges without problems, but two correlation parameters are estimated as 1.

### Principal components analysis for _maxLMM_

```{r m0PCA}
summary(rePCA(m0))
```

The PCA results indicate two dimensions with no variability in the random
effects for session and another two dimensions in the random effects for item.

### Zero-correlation-parameter linear mixed model (_zcpLMM_)

```{r gbm02}
m1 <- 
  lmer(sottrunc2 ~ 1+T+P+F+TP+TF+PF+TPF + (1+T+F+TF||session) + (1+T+P+TP||item),
       gb12, REML=FALSE)
VarCorr(m1)
anova(m1, m0)
```

The _zcpLMM_ fits significantly worse than the _maxLMM_, but it reveals several variance components with values close to or of zero.  

### Iterative reduction of model complexity

Let's refit the model without small variance components. 

```{r m2}
m2 <-
  lmer(sottrunc2 ~ 1+T+P+F+TP+TF+PF+TPF + (1+T+F||session) + (1+T||item), 
       gb12, REML=FALSE)
VarCorr(m2)
anova(m2, m1, m0)
```

Let's check the support of item-related variance components
```{r m3}
m3 <-
  lmer(sottrunc2 ~ 1+T+P+F+TP+TF+PF+TPF + (1+T+F||session) + (1|item),
       gb12, REML=FALSE)
VarCorr(m3)
anova(m3, m2)
```

Marginally significant drop. (Deleting also intercept leads to significant drop in goodness of fit.)

### Extending the reduced LMM with correlation parameters

Let's check correlation parameters for `SubjID`

```{r m4}
m4 <- 
  lmer(sottrunc2 ~ 1+T+P+F+TP+TF+PF+TPF + (1+T+F|session) + (1|item), 
       gb12, REML=FALSE)
VarCorr(m4)
anova(m3, m4)
```

The correlation parameter is significant, but one correlation is 1.000, indicating a singular model. Let's remove the small correlation parameters.

### Pruning small correlation parameters

```{r m5}
m5 <-
  lmer(sottrunc2 ~ 1+T+P+F+TP+TF+PF+TPF + (1+F|session) + (0+T|session) + (1|item),
       gb12, REML=FALSE)
VarCorr(m5)
anova(m5, m4)
```

Now the model is clearly degenerate: The correlation is at the boundary (-1); `theta` returns a zero value for one of the variance components.
 
## Summary


## Versions of packages used
```{r versions}
sessionInfo()
```

